mimix: Reference-based imputation of missing data

### 2021-03-03 paper/paperRmimix.tex
Comments on manual and updates to paper based on manual

% software 
The RBI methods of \citet{Carpenter++13} were implemented in SAS in what have become known as ``the five macros'' and are available on the web page (on www.missingdata.org.uk) of the DIA working group for missing data. These macros also provide for ``delta-adjustment'' in which imputed values are modified by a user-specified amount \citep{Ratitch++13}: this is useful is performing sensitivity analysis, since the RBI methods make a number of untestable assumptions \citep{ian:ZAMSTAR}.
The methods were then implemented in Stata by \citet{Cro++16}.

@@

CRK proposed corresponding variance matrices. Following \cite{ian:RBIcausal}, we describe only the regression coefficient matrix and conditional variance matrix of the potential outcomes after visit  $t$ given those before visit $t$. 
CRK set these to be taken from arm $z$ for MAR and LMCF, 
and from arm $r$ for J2R, CIR and CR.

[KEVIN: An approach that we call \emph{RBI alternative} instead uses
$\boldsymbol{\beta}_t(\tmax)$ and $\boldsymbol{\Omega}_t(\tmax)$ for all RBI methods: did we code this???]

@@

[KEVIN - can we let $(k_0, k_1)$ depend on arm?]

@@

We follow the approach of the five macros. Any imputed value $Y^*_{z,t}$ is replaced by $Y^*_{z,t} + \sum_{u=t+1}^{u=\tmax} a_u b_{u-t}$ 
where $a_u$ is a user-specified shift for time $u$ and $b_w$ is a user-specified scaling multiplier that controls how the user-defined shift for time $u$ is applied to all times $\ge u$. For example, if $b=(1,0.7, 0.5,...)$ then a participant who discontinues treatment just after visit $t$ has 
their imputed value at time $t+1$ incremented by $a_{t+1}$, 
their imputed value at time $t+2$ incremented by $0.7 a_{t+1} + a_{t+2}$, 
their imputed value at time $t+3$ incremented by $0.5 a_{t+1} +0.7  a_{t+2} + a_{t+3}$, and so on. 
Delta adjustment can apply either with RBI or the causal method and is always applied after imputation.

\subsection{Interim missing values and covariates}

[Kevin - to discuss]

@@

KEVIN TO DRAFT - INITIALLY TAKE FROM HELP.

\subsection{Arguments}
\subsection{Outputs}

@@

KEVIN TO DRAFT. Which data to use?

@@

\begin{itemize}
\item flexible package
\item doesn't cover other outcome types; e.g. methods have been proposed for recurrent events \citep{Keene++14}
\item trials should be designed consistently with their estimand, so if the treatment-policy estimand is of interest then outcomes should be collected after treatment discontinuation. Analysis options for this setting are still in development: options include including treatment discontinuation time in the model and imputing under MAR [ref Tom at GSK]; using RBI or causal model and reserving the post-discontinuation data for model checking; or using RBI or causal model and using the post-discontinuation data to estimate model parameters such as $K_{z,t}$. 
\end{itemize}

>>>>

% software 
The RBI methods of \citet{Carpenter++13} were implemented in SAS in what have become known as ``the five macros'' and are available on the web page (on www.missingdata.org.uk) of the DIA working group for missing data. These macros also provide for ``delta adjustment'' in which imputed values are modified by a user-specified amount \citep{Ratitch++13}: this is useful is performing sensitivity analysis, since the RBI methods make a number of untestable assumptions \citep{ian:ZAMSTAR}.
The methods were then implemented in Stata by \citet{Cro++16}.
A practical guide to their use is given by \citet{Cro++20}.

@@

The variance matrices are constructed so that the regression coefficient matrix and conditional variance matrix of the potential outcomes after visit  $t$ given those before visit $t$ are taken from arm $z$ for MAR and LMCF, and from arm $r$ for J2R, CIR and CR 
\citet{Carpenter++13,ian:RBIcausal}.
[NOTE we didn't code the \emph{RBI alternative} approach of \citet{ian:RBIcausal} that instead uses $\boldsymbol{\beta}_t(\tmax)$ and $\boldsymbol{\Omega}_t(\tmax)$ for all RBI methods.]

@@

[AT PRESENT $(k_0, k_1)$ ARE FIXED BUT IT MAY BE POSSIBLE TO LET THEM DEPEND ON ARM. THIS MAKES SENSE FOR A TRIAL WITH MORE THAN 2 ARMS. WOULD PROBABLY CODE THEM AS A VARIABLE IN DATA SET SO COULD LET THEM VARY BY OTHER FACTORS E.G. REASON?]

@@

Delta adjustment allows imputations to differ sytematically from RBI methods and so provides a framework for performing sensitivity analysis around RBI assumptions.
We follow the approach of the five macros. 
Delta adjustment provides an increment which is added on to all values imputed after treatment discontinuation, but not to interim missing values. 
It is specified by the options delta and dlag which we denote as $a_u$, a user-specified shift for time $u$, and $b_w$, a user-specified scaling multiplier that controls how the user-defined shift for time $u$ is applied to all times $\ge u$.
Formally, any imputed value $Y^*_{z,t}$ is replaced by $Y^*_{z,t} + \sum_{u=t+1}^{u=\tmax} a_u b_{u-t}$.
This means that values of delta are cumulated after treatment discontinuation.

For example, consider an individual who discontinued treatment at the 2nd time point. 
Under the setting \texttt{dlag=c(1,1,...,1)}, we take the vector of delta’s starting at the 3rd time point and add their cumulative sums to the imputed values. 
Modifying dlag modifies this behaviour, so that the vector of delta’s starting at the 3rd time point is multiplied elementwise by the vector dlag. 
A common increment of 3 at all time points after treatment discontinuation would be achieved by setting \texttt{delta=c(3,3,3,...)} and \texttt{dlag=c(1,0,0,...)}.
An increment of 3 at the time point immediately after discontinuation that is halved at each subsequent time point is specified by  \texttt{delta=c(3,3,3,...)} and \texttt{dlag=c(1,-1/2,-1/4,...)}.

%For example, if $b=(1,0.7, 0.5,...)$ then a participant who discontinues treatment just after visit $t$ has 
%their imputed value at time $t+1$ incremented by $a_{t+1}$, 
%their imputed value at time $t+2$ incremented by $0.7 a_{t+1} + a_{t+2}$, 
%their imputed value at time $t+3$ incremented by $0.5 a_{t+1} +0.7  a_{t+2} + a_{t+3}$, and so on.

Delta adjustment can apply either with RBI or the causal method and is always applied after imputation.

\subsection{Interim missing values}

Interim missing values are values that are missing while the individual remains on treatment and are identified by being followed by later observed values. They are always imputed under MAR. Imputing interim missing values in a different way from post-discontinuation missing values is the procedure that is most likely to be appropriate in practice, since interim missing values that occur while on treatment are unlikely to follow the same pattern as post-discontinuation missing values that occur when off treatment. This differs from the behaviour of the Stata package \citep{Cro++16} which by default imputes interim missing values using the same procedure as post-discontinuation values, with MAR as an option.

\subsection{Covariates}

Covariates are handled ... [HOW? IAN THINK]

@@

\subsection{Arguments} [KEVIN ARE WE ALLOWED TO CHANGE THE ORDER? BELOW MAKES MOST SENSE TO ME.]

\begin{tabular}{llp{.8\linewidth}}\hline
& Option & Description \\ \hline
\multicolumn{3}{l}{\em Data options} \\
& data 	&	 Dataset in long format, i.e.\ with one record per individual per time point. \\
& covar 	&	 Covariates measured at or before baseline. Must be complete (no missing values).	\\
& depvar 	&	 Dependent (outcome) variable.	\\
& treatvar 	&	 Treatment group, coded 1,2,...	\\
& idvar 	&	 Participant identifier.	\\
& timevar 	&	 Time point for repeated measures. \\

\multicolumn{3}{l}{\em Method options} \\
& method 	&	 Reference-based imputation method: must be one of J2R, CIR, CR, LMCF, MAR, Causal.	\\
& methodvar 	&	 Alternative to method option: variable in dataset specifying the imputation method  for each individual. \\
& reference 	&	 Reference group for J2R, CIR, CR and Causal methods. \\
& referencevar 	&	 Alternative to reference option: variable in dataset specifying the reference group for each individual. \\

\multicolumn{3}{l}{\em Causal method options} \\
& K0 	&	 Causal constant for use with Causal method. The treatment effect assumed at post-discontinuation times is K0 times the treatment effect at the time of discontinuation; if K1 $\ne$ 1, this is multiplied by a decaying term. \\
& K1 	&	 Causal constant for use with Causal method. The treatment effect assumed at post-discontinuation times, implied by K0, decays exponentially by a factor K1 per time unit. \\

\multicolumn{3}{l}{\em Delta method options}\\
& delta 	&	 Optional vector of delta values to add onto imputed values (a’s in Rogers paper). Length must equal number of time points.\\
& dlag 	&	 vector of delta values to add onto imputed values (non-mandatory) (b’s in Rogers paper) Length must equal number of time points. Default is c(1,1,1,...). \\

\multicolumn{3}{l}{\em Computation options}\\
& M 	&	 Number of imputations to be created. \\
& seed 	&	 Seed value. Specify this so that a new run of the command will give the same imputed values.	\\

\multicolumn{3}{l}{\em MCMC options} \\
& prior 	&	 Prior for the variance-covariance matrix when fitting multivariate normal distributions: Jeffreys (default), uniform or ridge.	\\
& burnin 	&	 Number of burn-in iterations when fitting multivariate normal distributions.	\\
& bbetween 	&	 Number of iterations between imputed data sets when fitting multivariate normal distributions.	\\
& mle 	&	 Logical option to use maximum likelihood parameter estimates instead of MCMC draw parameters. [NEED TO WRITE A COMMENT ABOUT THIS -- OR UNDOCUMENT IT?]	\\ \hline
\end{tabular}

\subsection{Algorithm}

The program works through the following steps.
\begin{enumerate}
\item Set up a summary table based on treatment arm and missing data pattern (i.e.\ which timepoints are unobserved).
\item Fit a multivariate normal distribution to each treatment arm using MCMC methods in package norm2 \citep{norm2}.
\item Impute all interim missing values under a MAR assumption, looping over treatments and patterns.
\item Impute post-discontinuation missing values under the user-specified assumption, looping over treatments and patterns. If methodvar and/or refvar is specified then the loop is also over teh values of these.
\item Perform delta adjustment if specified.
\item Repeat steps 2-5 M times and form into a single data frame.
\end{enumerate}
The baseline value of the outcome could be handled as an outcome, but this would allow a treatment effect at baseline [IAN CHECK]. We instead recommend handling it as a covariate.

The program is based on Suzie Cro’s Stata program mimix.

\subsection{Outputs}

The M imputed data sets are output concatenated as one large dataset appended to
the original unimputed dataset. [KEVIN IS IT A DATA SET, A DATA FRAME, OR OTHER?]

The user can use the as.mids() function in the mice package to convert the output data  to mids data type and hence to perform analysis using Rubin's rules.

@@

KEVIN TO DRAFT PLEASE USING ASTHMA DATA.

@@

\begin{itemize}
\item flexible package
\item doesn't cover other outcome types; e.g. methods have been proposed for recurrent events \citep{Keene++14}
\item The suitability of the Rubin's rules standard errors has been debated \cite{ian:Carpenter_letter,Carpenter++14,Cro++19}...
\item trials should be designed consistently with their estimand, so if the treatment-policy estimand is of interest then outcomes should be collected after treatment discontinuation. Analysis options for this setting are still in development: options include including treatment discontinuation time in the model and imputing under MAR [ref Tom at GSK]; using RBI or causal model and reserving the post-discontinuation data for model checking; or using RBI or causal model and using the post-discontinuation data to estimate model parameters such as $K_{z,t}$. 
\end{itemize}
